Behavioural Analysis of Independent Value-Based Learning in Non-cooperative Games

### 2025-06-04 paper/paper.tex
paper v6

\documentclass[a4paper]{article}

% Packages
\usepackage{geometry}
\geometry{left=1.5cm, right=1.5cm, top=2.54cm, bottom=2.54cm}
\usepackage{graphicx, hyperref, setspace, amsmath, amssymb, titlesec, multicol, parskip, indentfirst, etoolbox, caption, xcolor, svg}
\usepackage[style=ieee]{biblatex}
% Import the bibliography file
\addbibresource{ref.bib}

% Title Formatting
\titleformat{\section}{\centering\large\scshape}{\thesection}{1em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection.}{1em}{}
\setstretch{1.0} % Keep single spacing
\setlength{\parskip}{6pt} % Space between paragraphs
\titlespacing{\section}{0pt}{6pt}{6pt}
\titlespacing{\subsection}{0pt}{6pt}{6pt}
\titlespacing{\subsubsection}{0pt}{6pt}{6pt}
% Document Title
\title{
    \textbf{Draft v1}
}

\date{}
\captionsetup{labelfont={small,sc}, textfont={small,sc}}
% Section Numbering
% Define numbering format
\renewcommand{\thesection}{\Roman{section}.}
\renewcommand{\thesubsection}{\textit{\Alph{subsection}.}}
\renewcommand{\thesubsubsection}{\textit{\arabic{subsubsection}.}}
\renewcommand{\thetable}{\Roman{table}} % Set table numbering to Roman
\renewcommand{\thefigure}{\Roman{figure}} % Number figures in Roman numerals

% Make subsection titles italic as well
\titleformat{\subsection}{\normalfont\large\itshape}{\thesubsection}{1em}{}
\titleformat{\subsubsection}{\normalfont\itshape}{\thesubsubsection}{1em}{}

@@

\maketitle
\vspace{-1.5cm}

% Authors Block
\begin{multicols}{3}
    \centering
    \textbf{Yang Li}\\
    \textit{University of the West of England}\\
    \textit{yang4.li@uwe.ac.uk}\\
    \vfill

    \textbf{Marco Perez Hernandez}\\
    \textit{University of the West of England}\\
    \textit{marco.perezhernandez@uwe.ac.uk}\\
    \vfill

    \textbf{Mehmet Aydin}\\
    \textit{University of the West of England}\\
    \textit{mehmet.aydin@uwe.ac.uk}\\
    \vfill
\end{multicols}

\singlespacing
\setlength{\parskip}{6pt}
\setlength{\parindent}{0.5cm}

\begin{abstract}
The abstract should be written after the rest of the paper is completed.
\end{abstract}

\begin{multicols}{2}
\setlength{\columnsep}{0.5cm}

\section{Introduction}
Many real-world problems can be modeled as non-cooperative games, such as business competition, political campaigns, stock trading, bidding in auctions, cyber security, and traffic routing. These problems feature the competitive nature of games because they share one thing in common, parties with conflicting interests. Plenty of research has been done on multi-agent reinforcement learning (MARL) in fully cooperative games. There is currently a gap in the non-cooperative setting \cite{zhu2024survey}. Most current SOTA MARL algorithms are designed for cooperative games.

In fully cooperative games, there is no conflict between agents. If agents can communicate, they are motivated to form a firm alliance that essentially works as a collective mind that is the one and only conscious being. Thus, in a game theory sense, a fully cooperative game where agents share information is equivalent to a single agent game where the collective agent tries to find the optimal joint action, given the joint observation. To do that in a traditional single agent setting where the environment is stationary and non-adversarial, one only needs to learn the environment mechanics (the state transition function) and the reward function so that the Q-values of an action in a given state can be computed or predicted by a function approximator. A Pareto optimal pure (deterministic) strategy is guaranteed to exist in fully cooperative finite games, meaning that the agents can always deterministically choose the optimal action that has the highest payoff. The only thing that keeps the agents from forming a cluster in practice is the hardware incapability (computational power, real-world constraints, etc.).

Non-cooperative environments, however, involve conflicts that make learning more complicated. Agents may belong to different parties that would like to have private control to protect their own interests. For example, the stock market is often considered a 0-sum fully competitive game where one person's profit is another person's loss, if we omit the long-term economic growth.

For cooperative games, we can always say that 5 rewards is better than 2 rewards, but what about the rewards $\langle 1,5 \rangle$ and  $\langle 10,1 \rangle$ where each agent gets one value in the vector? Rational self-interested agents will have to fight for what is best for them. Many solutions in game theory are possible. One of the commonly accepted solutions for is the Nash equilibrium (NE). NEs are normally what we want the agents to converge to because convergence means stability and strict NEs are the only stable points in the game. Generally, there may only be a mixed NE, meaning that players adopt randomized strategies consisting of probability distributions over the action space.

Currently, most SOTA MARL algorithms cannot learn a mixed policy. Largely due to the studied games are mostly fully cooperative or asynchronous. As we will show in the experiments, those algorithms will fail to find the optimal action in games specifically designed to challenge the agents' competitive thinking. We need to reconsider our approach if we want to build agents that are capable of dealing with non-fully-cooperative environments.

\subsection{Aim}
The aim of this study is to point out that current SOTA algorithms can be misleadingly performing well in non-fully-cooperative environments by both formal analysis and showing specific game scenarios that are designed to challenge the agents from a game theory perspective.

\subsection{Stochastic Game}

A stochastic game is formally defined by a tuple:

$G = \langle I,S,S_0,S_t,\{A_1,\ldots,A_n\},\{R_1,\ldots,R_n\},T \rangle$
\begin{itemize}
    \item $I$: the set of agents, from 1 to \(n\).
    \item $S$: the game’s state space.
    \item $S_0$: the set of possible initial states.
    \item $S_t$: the set of terminal states.
    \item $\{A_1,\ldots,A_n\}$: one action space for each agent.

    Let $a_i \in A_i$ be an action of agent $i$, and $a^n=\langle a_1,\ldots,a_n \rangle \in A^n$ denote a joint action in the joint action space.
    \item $\{R_1,\ldots,R_n\}$: one reward function for each agent.

    $R_i:S \times A^n \times S \rightarrow \mathbb{R}$
    \item $T$: the state transition function.

    $T:S \times A^n \times S \rightarrow [0,1]$

    It defines the probability of the game transitions from one state to another given the joint action. It must satisfy the axiom of probability:
    \begin{equation}
    \forall s \in S,a^n \in A^n: \sum_{s' \in S}{T(s,a^n,s')}=1
    \label{eq:statetransition}
    \end{equation}
    It is convenient to define $T^*:S \times A^n \rightarrow \Pi_S$, where $\Pi_S$ denotes the set of all probability distribution over the set S.

    Some games incorporate partial observability and define an observations space, which is irrelevant to the purpose of this study.
\end{itemize}

 Starting from a randomly chosen $s_0 \in S_0$, at each timestep $t \in \mathbb{N}$ with the current state $s_t$, each agent observes the current state $s_t$ and chooses an action forming a joint action $a^n=\langle a_1,\ldots,a_n \rangle$. The game then moves into the next state $s_{t+1}$, randomly sampled from the distribution P and a reward $R_i$ is assigned to each agent. If $s_t \in S_t$, the game ends. Otherwise, the game continues in the next timestep.

The goal is to find the optimal strategy (policy) for an agent to pick actions given the current state information they receive from the environment, such that their expected cumulative reward is maximized.

To demonstrate the problem with competitive games, let's analyze some simple normal form games. A normal form game has only two states, the starting state and the end state. The game ends immediately after all agents pick an action. Thus, they can be represented using the reward matrix. Consider rock-paper-scissors (simple one-shot matrix game) in Table~\ref{table:rps} for example. Each entry represents the reward for the row player and the column player, respectively.

\noindent
\begin{minipage}{\columnwidth}
\captionof{table}{Rock Paper Scissors} 
\label{table:rps}
\centering
\begin{tabular}{c|ccc}
    & r & p & s\\
    \hline
    r & 0,0 & -1,1 & 1,-1 \\
    p & 1,-1 & 0,0 & -1,1 \\
    s & -1,1 & 1,-1 & 0,0 \\
\end{tabular}
\end{minipage}

The only NE in the game is that two players play each of the three actions with 1/3 chance. If any player deviates from the NE, their expected reward does not increase, and the other player can exploit them by playing the action that counters their most frequent action. Here is a more complicated game called biased rock paper scissors.

\noindent
\begin{minipage}{\columnwidth}
\captionof{table}{Biased Rock Paper Scissors} 
\label{table:biasedrps}
\centering
\begin{tabular}{c|ccc}
    & r & p & s\\
    \hline
    r & 50,50 & 25,75 & 100,0 \\
    p & 75,25 & 50,50 & 45,55 \\
    s & 0,100 & 55,45 & 50,50 \\
\end{tabular}
\end{minipage}

The only NE in the game is that both players adopt the following policy.
\begin{equation}
\pi=(P(r)=0.0625,P(p)=0.6250,P(s)=0.3125)
\label{eq:biasedrpspolicy}
\end{equation}

The RL algorithm needs to learn $\pi$, not the optimal action because there is no optimal action. For competitive play after the learning process, the agent should play randomly according to the learned distribution.

\section{MARL in Competitive Games}
Algorithms that are carried over from the traditional single agent RL domain, such as Q-learning, do not incorporate randomized policy, even though they may behave randomly during the learning process (using an $\epsilon$-greedy exploration vs. exploitation strategy, for example). In the single agent setting, the agent only needs to learn the expected long-term return (the Q-value) of each action and deterministically pick the highest. Thus, they cannot learn $\pi$.

There is a family of algorithms called Joint Action Learning (JAL) that addresses the problem by incorporating the Q-values for the joint action. Minimax Q-learning \cite{littman1994markov} is a JAL algorithm that can be applied to two-player zero-sum games. It computes local minimax solutions for each state based on the learned joint Q-values. Nash Q-learning \cite{hu2003nash} is a similar algorithm that computes the local NE instead. It converges to a NE under highly strict assumptions that is found in almost no games. Correlated Q-learning \cite{greenwald2003correlated} computes a correlated equilibrium and has no known convergence guarantees. \cite{zinkevich2005cyclic} showed that JAL cannot converge in certain types of ill-conditioned games.

Some other research \cite{qu2020distributed} tried to address this problem using a manually calculated distribution over the actions based on the payoff, such that the higher the payoff of an action, the more likely the action will be chosen. This approach intuitively makes sense. The problem is that in any NE, for any action that are chosen with a positive probability, the payoff of the actions are the same. If we calculate the expected reward of each action in the biased rock-paper-scissors example where the opponent uses the NE strategy, we will see all three actions have the same payoff.

Suppose the algorithm can converge to a NE, then all Q-values should be the same (ignoring worthless actions). Such an algorithm will output a uniform distribution. The algorithm is essentially learning a pure random policy. This contradicts with the assumption that they converge to a NE. We have shown that approach cannot work through a proof by contradiction.

Most MARL problems studied are much more than single state matrix games and have astronomically large state/action spaces. It is impossible for the agents to converge to a global NE in the first place because of the limited computational power. Because of the complexity of the environment that many studies carry out in, picking the action with the best Q-value still gives reasonably good results since at least the agents learned to interact with the environment. The competitiveness of the game is not going to play a huge role before the agents can really predict the state transitions. Currently, many fully competitive games are studied in the MARL community such as Pommerman \cite{resnick2018pommerman}, Go \cite{silver2017mastering}, StarCraft II \cite{vinyals2019grandmaster}, and Dota 2 \cite{berner2019dota}. There are also some general environments that are non-cooperative such as hide-and-seek \cite{baker2019emergent}, Neural MMO \cite{suarez2021neural}, PettingZoo \cite{terry2021pettingzoo}, and Melting Pot \cite{leibo2021scalable}. However, we lack a standard benchmark for MARL algorithms in non-cooperative games due to the difficulty of measuring agents' performance in general non-cooperative games.
FightLadder \cite{li2024fightladder} used the Elo rating system to test MARL algorithms but they were restricted to two-player zero-sum fighting games.

\section{Methods}
We define the mechanics of the game Food Chain. I also will have to migrate the tables and charts from my previous Word documents to explain the unique characteristics that make it challenging which are not found in other games (especially compare to the hide and seek game).

The game is complex enough to eliminate any tabular approaches. The state space is roughly 1060 for just a 9x9 board and can grow exponentially. Function approximation methods, such as deep neural networks or Monte Carlo methos must be used to solve the whole game. The spacial information on the game board can also be a good challenge to algorithms using CNNs. The game is still simple enough to be human playable. The mechanics of the game (the state transition function) is very straightforward and easy for the agents to learn, so that they can focus on agent interactions. Because of the large state space, it is impossible for us to check whether the agents converged to a global NE since it is NP-Hard. Yet it has carefully designed mechanics (see my Casual Monthly Progress Report for Nov 2024) which very often force the agents to adopt random policies (and addressing the coordination problem, too). These are the leaf nodes of the game tree, which is equivalent to a simple matrix game. We can actually check whether the agents converge to NEs in those sub-cases easily. So, in summary.
\begin{itemize}
    \item It is simple enough to let the agents focus on interactions instead of the environment.
    \item It is complex enough to be interesting.
    \item It forces the agents to constantly address the mixed strategy problem.
    \item We can analyze the performance of the agents with game theory.
\end{itemize}

\section{Results}
Figure \ref{fig:hopperreward}, \ref{fig:frogreward}, \ref{fig:snakereward} shows the average test return for each type of agent after training in a randomly generated game environment. Figure \ref{fig:scenhopperreward}, \ref{fig:scenfrogreward} demonstrates that in the carefully set up matching pennies scenario. The results show that the algorithms could not reach the optimal reward which can be computed using linear algebra (which is also another strong aspect of the game. we can analyze the game formally and see whether the agents reached optimality). More formal analysis coming...

\noindent
\begin{minipage}{\columnwidth}
\centering
\includesvg[width=1\textwidth]{plots/rewards_global_hopper.svg}
\captionof{figure}{hopper rewards}
\label{fig:hopperreward}
\end{minipage}

\noindent
\begin{minipage}{\columnwidth}
\centering
\includesvg[width=1\textwidth]{plots/rewards_global_frog.svg}
\captionof{figure}{frog rewards}
\label{fig:frogreward}
\end{minipage}

\noindent
\begin{minipage}{\columnwidth}
\centering
\includesvg[width=1\textwidth]{plots/rewards_global_snake.svg}
\captionof{figure}{snake rewards}
\label{fig:snakereward}
\end{minipage}

\noindent
\begin{minipage}{\columnwidth}
\centering
\includesvg[width=1\textwidth]{plots/rewards_scen_hopper.svg}
\captionof{figure}{snake rewards}
\label{fig:scenhopperreward}
\end{minipage}

\noindent
\begin{minipage}{\columnwidth}
\centering
\includesvg[width=1\textwidth]{plots/rewards_scen_frog.svg}
\captionof{figure}{snake rewards}
\label{fig:scenfrogreward}
\end{minipage}

\section{Discussion}

\section{Conclusion}

>>>>

\documentclass[]{interact}

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage[caption=false]{subfig}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
%\usepackage[doublespacing]{setspace}% To produce a `double spaced' document if required
%\setlength\parindent{24pt}% To increase paragraph indentation when line spacing is doubled
\usepackage{svg}

\usepackage[numbers,sort&compress]{natbib}% Citation support using natbib.sty
\bibpunct[, ]{[}{]}{,}{n}{,}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
\makeatletter% @ becomes a letter
\def\NAT@def@citea{\def\@citea{\NAT@separator}}% Suppress spaces between citations using natbib.sty
\makeatother% @ becomes a symbol again

\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

@@

%\articletype{ARTICLE TEMPLATE} Specify the article type or omit as appropriate

\title{Behavioural Analysis of Independent Value-Based Learning in Non-cooperative Games}

\author{
\name{Yang Li\textsuperscript{a}\thanks{CONTACT Yang Li. Email: yang4.li@uwe.ac.uk}, Marco Perez Hernandez\textsuperscript{a} and Mehmet Emin Aydin\textsuperscript{a}}
\affil{\textsuperscript{a} University of the West of England, Coldharbour Lane, Frenchay Campus, Bristol, UK}
}

\maketitle

\begin{abstract}
Multi-agent reinforcement learning has received a lot of updates in cooperative games recently. However, research in general non-cooperative games is currently lagging behind. Value-based algorithms such as Q-learning have been one of the most widely studied methods in reinforcement learning, and their independent variations for multi-agent environments are popular approaches due to their simplicity and good scalability. Some research has shown that independent algorithms can heavily outperform other approaches in some cooperative scenarios. In this paper, we discuss the feasibility of these algorithms in non-cooperative settings. As a result, we explain the conditions a game must satisfy for the algorithms to work. We further propose the game Food Chain. The game simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our formal analysis. Our results show that independent value-based learning algorithms are guaranteed to fail in some non-cooperative games.
\end{abstract}

\begin{keywords}
Multi-agent; reinforcement learning; independent value-based learning; game theory; non-cooperative games; competitive games; Nash equilibrium
\end{keywords}

\section{Introduction}
Recent multi-agent reinforcement learning (MARL) research has mostly focused on fully cooperative games \cite{yu2020benchmarking, papoudakis2020benchmarking, zhu2024survey} where there is no conflict between agents as they share a common reward. If we are allowed to share information among the agents, a fully cooperative game reduces to a single-agent game where the collective agent tries to find the optimal joint action, given the joint observation, although the problem is still challenging in practice because of the computational complexity of the joint action brings. In a traditional single-agent setting, the agent needs to learn which action gives the best reward. A Pareto-optimal pure strategy is guaranteed to exist in fully cooperative finite games, meaning that the agents can always deterministically choose the optimal action that has the highest payoff.

Non-cooperative environments, however, involve conflicts that make learning more complicated. Agents may belong to different parties that would like to have private control to protect their own interests. Rational self-interested agents will have to fight for what is best for them, as increasing one agent's reward might lead to the decline of other's gains. We can no longer intuitively define the solution for a game as the optimal strategy resulting in the highest reward. One of the possible solution concepts is the Nash Equilibrium (NE). NEs describe the stable points in games, which are associated with the convergence to optimal behaviour of agents.

Many real-world problems can be modelled as non-cooperative games, such as business competition, political campaigns, stock trading, bidding in auctions, cyber security, and traffic routing. For example, if we omit the long-term economic growth, the stock market is often considered a 0-sum fully competitive game where one person's profit is another person's loss. Another example is cyber security, which must study the conflicts between attackers and defenders. These problems feature the competitive nature of games because they share one thing in common, parties with conflicting interests. Substantial research has been done on MARL in fully cooperative games, and non-cooperative games have received less attention \cite{zhu2024survey, li2024fightladder}.

There are many attempts to conquer games with MARL. One of which is independent value-based learning (IVBL). IVBL is a family of algorithms that models the multi-agent environments as single-agent environments, and they compute for each agent the expected reward of each action under each state. IVBL is simple and highly scalable because it is decentralized. It has been adopted for both cooperative \cite{foerster2017stabilising, omidshafiei2017deep, palmer2017lenient, palmer2018negative} and non-cooperative tasks \cite{bjornsson2009cadiaplayer, jiang2018q, qu2020distributed, kopacz2023evaluating}. This paper analyses the performance of IVBL in non-cooperative finite games. The analysis considers characteristics of the action space and the NEs of the game, combining both formal analysis and experimental evaluation in selected two-player non-cooperative games. The results show NEs with uniformly random policies is a necessary condition for stability convergence of IVBL algorithms in non-cooperative environments. These condition highly restricts the wider applicability of IVBL.

\subsection{Main contributions}
\begin{itemize}
    \item Formal analysis identifying necessary conditions for stability convergence of IVBL which can drive the choice of algorithms in non-cooperative finite games.
    \item Introducing a non-cooperative game that has the characteristics which challenge the agents from a game theory perspective aligned with our formal analysis.
    \item Experimental evaluation of IVBL algorithms in selected two-player scenarios in the game proposed above.
\end{itemize}

\subsection{Contents}
The remainder of the paper is organized as follows:
In the next section, we provide the literature review on the key features of non-cooperative games and the most popular MARL algorithms that is related to our work. We then state the formal definitions of games and other concepts discussed in this paper in section 3. Section 4 analyses the stability of IVBL in non-cooperative games. To verify our analysis with experimental results, a new game called Food Chain is introduced in section 5. The game will have the key features discussed in our analysis that bring specific challenges to MARL agents. Then, we present the design as well as the results of our experiments in section 6. Finally, draw our conclusion in the last section.

\section{Tackling Non-Cooperative Games}
This literature review focuses on key features of non-cooperative games, as well as the MARL approaches to solve these games.

\subsection{Non-cooperative Games}
In single-agent or fully cooperative environments, there is one reward function for all agents. Thus, the agents only need to learn which (joint) action has the highest payoff. In non-cooperative games, there may be no optimal joint action that satisfy all agents. Many solution concepts to non-cooperative games are discussed in game theory \cite{albrecht2024multi}, such as Pareto optimum, correlated equilibrium, and Nash equilibrium. John Nash is famous for his study on non-cooperative games and his concept of Nash Equilibrium (NE) \cite{nash2024non}. NE describes a stable situation where no player can benefit from deviating from their current strategy independently. NEs offer a framework to characterise and analyse stability in games. Stability is associated to the optimal outcome in non-cooperative games and hence the objective agents aim for. Generally, there may only be a mixed NE, meaning that players must adopt randomized action-selection policy consisting of probability distributions over the action space. This is the mixed policy problem.

Many fully competitive games are studied in the MARL community such as Pommerman \cite{resnick2018pommerman}, Go \cite{silver2017mastering}, StarCraft II \cite{vinyals2019grandmaster}, and Dota 2 \cite{berner2019dota}. There are also some general non-cooperative environments such as hide-and-seek \cite{baker2019emergent}, Neural MMO \cite{suarez2021neural}. PettingZoo and Melting Pot \cite{terry2021pettingzoo, leibo2021scalable} are two integrated MARL environments including many cooperative and non-cooperative games. \cite{kopacz2023evaluating} studied the non-cooperative game predator-and-prey in the PettingZoo library, which is closely related to the game defined in our work. However, we lack a standard benchmark for MARL algorithms in non-cooperative games due to the difficulty of measuring agents' performance in general non-cooperative games. In cooperative games, the reward will simply increase as the agents get better at the game. This is no longer the case in non-cooperative games where increasing one agent's outcome often results in the drop of another agent's return. FightLadder \cite{li2024fightladder} used the Elo rating system to test MARL algorithms but they were restricted to two-player zero-sum fighting games.

\subsection{MARL Algorithms}
Independent Learning (IL) algorithms that are carried over from the traditional single-agent RL domain treat other agents as part of the environment. The agents are essentially unaware of the existence of each other. IL is popular for handling many complex problems \cite{papoudakis2020benchmarking, gupta2017cooperative, de2020independent, palmer2020independent}. Independent value-based learning (IVBL) algorithms, such as the classic independent Q-Learning (IQL) \cite{tan1993multi}, learn the expected long-term cumulative return (the Q-value) of each action at each state through sampling. IVBL algorithms including IQL are often considered a baseline and a natural choice in a wide range of MARL studies because of their simplicity and scalability. For example, they were studied in various cooperative environments \cite{foerster2017stabilising, omidshafiei2017deep, palmer2017lenient, palmer2018negative}. The problem is that IVBL agents only give us the average reward of each action for each agent, but do not tell us what the action-selection policy look like. This is not a problem in single-agent or cooperative environments as we can simply pick the action with the highest return. Some studies addressed this issue in non-cooperative games by simply choosing the action with the highest empirical return \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating} and observed reasonably good outcome. \cite{kopacz2023evaluating} has examined IVBL in the non-cooperative game predator-and-prey. Both the algorithm and the game are closely related to our work. Other researchers \cite{qu2020distributed} tried to address this problem using the softmax function (\ref{eq:softmax}) to calculate a policy based on the payoff. Softmax takes as input a vector and computes a probability distribution over the original vector elements, such that the higher the original value, the higher the probability.
\begin{equation}
    \textbf{softmax}([x_1, ..., x_n]) = [\frac{e^{x_1}}{\sum_{i}{e^{x_i}}}, ..., \frac{e^{x_n}}{\sum_{i}{e^{x_i}}}]
    \label{eq:softmax}
\end{equation}
The latter approach intuitively makes sense, as it emphasis the action with higher empirical return. However, these approaches have a huge problem that guarantees failure of learning in a wide range of games. We will take a closer look in the analysis section.

Unlike IL, joint action learning (JAL) algorithms allow the agent to gain access to the information processed by all other agents, such as their choices of actions, so agents can learn the expected return of joint actions. Minimax Q-learning \cite{littman1994markov} is a JAL algorithm that can be applied to two-player zero-sum games. It computes local minimax solutions for each state based on the learned joint Q-values. Nash Q-learning \cite{hu2003nash} is a similar algorithm that computes the local NE instead. It converges to a NE under highly strict assumptions that reduce the applicability of the approach to a limited set of games. Correlated Q-learning \cite{greenwald2003correlated} computes a correlated equilibrium and has no known convergence guarantees. \cite{zinkevich2005cyclic} showed that JAL cannot converge in certain types of ill-conditioned games. Centralized learning requires full knowledge of the action spaces of all agents, which may not always be available if we are dealing with real-world problems. Another major disadvantage is that the number of joint actions is the number of possible combinations of the individual actions. This introduces combinatorics to the computational complexity of our algorithms, making them unscalable.

In contrast to the value-based approach, policy-based algorithms, such as the policy gradient methods, work around this issue by learning the policies directly, without putting an estimation on the individual action return. Policy gradient methods optimize a parametrized policy function using gradient decent to maximize the return. An advantage of policy gradient methods is that they naturally work on continuous action spaces. The actor-critic algorithms are a family of RL algorithm which trains a parametrized policy called the actor and a state-value function called the critic to estimate the policy gradient. Soft actor-critic (SAC) \cite{haarnoja2018soft} has become one of the most popular algorithms recently. Its multi-agent variant (MASAC) has been considered the state-of-the-art (SOTA) baseline \cite{yu2020benchmarking, zhang2020lyapunov}. There is also naturally an independent version (ISAC) for MARL. The proximal policy optimization algorithm (PPO) \cite{schulman2017proximal} was based on actor-critic as well and got extended to multi-agent environments by some researchers \cite{yu2020benchmarking, yu2022surprising} (MAPPO). Surprisingly, the IL version of PPO (IPPO) has proven to substantially outperform MAPPO and other algorithms in some cooperative tasks in StarCraft II \cite{de2020independent}. MADDPG \cite{lowe2017multi} is another policy gradient algorithm that is often considered SOTA. Because it only works for continuous action spaces, we do not discuss them further in the context of finite games.

Other notable MARL algorithms that have shown good performance include VDN \cite{sunehag2017value} and its improved version QMIX \cite{rashid2020monotonic}. The weakness is that they can only work for fully cooperative games where agents share a common reward.

\subsection{Summary}
As shown in this review, the issue of deriving the policy from the Q-values learned by IVBL agents in non-cooperative games has not been fully addressed by the MARL research community, \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating, qu2020distributed} made progress in this direction, but they did not study the behaviour of the agents and identify the flaws of the algorithms from a game theory point of view.

\section{MARL Background}
In this section, we provide preliminaries including the formal definition of a game and NE.

\subsection{Stochastic Game}
A game is formally defined by a tuple:
$G = \langle I,S,S_0,S_t,\{A_1,\ldots,A_n\},\{R_1,\ldots,R_n\},T \rangle$
\begin{itemize}
    \item $N = \{1, ..., n\}$: the set of agents, numbered from 1 to $n$.
    \item $S$: the game’s state space.
    \item $S_0 \subset S$: the set of possible initial states.
    \item $S_t \subset S$: the set of terminal states.
    \item $\{A_1,\ldots,A_n\}$: one action space for each agent. Let $a_i \in A_i$ be an action of agent $i$, and $a^n=\langle a_1,\ldots,a_n \rangle \in A^n$ denote a joint action in the joint action space.
    \item $\{R_1,\ldots,R_n\}$: one reward function for each agent. $R_i:S \times A^n \times S \rightarrow \mathbb{R}$.
    \item $T$: the state transition function. $T:S \times A^n \times S \rightarrow [0,1]$. It defines the probability of the game transitions from one state to another given the joint action. It must satisfy the axiom of probability:
    \begin{equation}
        \forall s \in S,a^n \in A^n: \sum_{s' \in S}{T(s,a^n,s')}=1
        \label{eq:state_transition}
    \end{equation}
\end{itemize}
Starting from a randomly chosen $s_0 \in S_0$, at each time step $t \in \mathbb{N}$ with the current state $s_t$, each agent observes the current state $s_t$ and chooses an action forming a joint action $a^n=\langle a_1,\ldots,a_n \rangle$. The game then moves into the next state $s_{t+1}$, randomly sampled from the distribution P and a reward $R_i$ is assigned to each agent. If $s_t \in S_t$, the game ends. Otherwise, the game continues in the next time step.

The way a player plays the game defines a policy (strategy) as a function that computes the probability of selecting each action given the current state, $\pi:S \times A^n \rightarrow [0,1]$. A strategy is said to be pure if it is deterministic ($\pi:S \times A^n \rightarrow \{0,1\}$). Otherwise, we call it a mixed strategy. A set of strategies, one for each player, is called a profile $\pi^n$. Given the profile, we can compute the expected reward for each player in the game. The goal is to find the optimal strategy for an agent, or even the optimal profile, such that expected cumulative rewards are maximized. Cooperative games are special cases where agents share the same reward. The optimal profile is the one that maximizes the common reward. Because they share the same reward function, we can view the problem as a joint player selecting a joint action and receive a reward, which makes cooperative games essentially single-agent games from a game theory perspective. In general non-cooperative games, however, it may not be possible to make everyone happy as their are conflicting interests. A more profound definition of optimum need to be discussed in the non-cooperative games section.

Formally, let $R_i^*(\pi^n)$ be the function to compute the expected reward of the i'th player given the profile. Let $\pi_i$ be a policy of the i'th player and $\Pi_i$ be the policy space of the i'th player. Let $\pi^{-i}$ denote $\pi^n - \{\pi_i\}$, the set of strategies of all other players. A profile $\pi^n$ is a NE if
\begin{equation}
    \forall i \in N, \forall \pi_i \in \Pi_i: R_i^*(\{\pi_i\} \cup \pi^{-i}) \leqslant R_i^*(\pi^n)
    \label{eq:nash}
\end{equation}
A NE is called strict if $\leqslant$ is replaced by $<$. The idea is that NEs are the stable points in the game, because if an agent can benefit from switching strategy on its own, they would do so.

A game is finite if $|N|, |S|$ and all $|A_i|$ are finite, and the game always ends in a finite number of turns. John Nash proved that every finite game has at least one NE. To solve a game in the context of this paper means to find at least one NE. Games that are not finite can be ill-conditioned and have no NE. For example, a game where the player who names the greatest number wins has no best strategy. For the rest of the paper, we only refer to finite games. A normal form game has only two states, the starting state $S_0$ and the end state $S_1$. The game ends immediately after all agents pick an action. Thus, they can be represented using the reward matrix. In the book on game theory originally published in 1944 \cite{von2007theory}, Von Neumann and Morgenstern showed us that finite games can be translated into normal form games. If we encapsulate the following states as a subgame, we end up with a reward matrix for the initial state with the rewards being the expected rewards of the subgames. We focus on normal form games because they are easier to work with for the formal analysis. Since they only have one state $S_0$ for agents to take actions, we simply write $\pi(a)$ instead of $\pi(S_0, a)$ to denote the probability of taking action $a$.

\section{Stability of IVBL in Non-Cooperative Games}
In this part of the paper, we show the feasibility of IVBL algorithms in non-cooperative games with formal analysis. We demonstrate the necessary conditions the game must satisfy in order to apply the IVBL algorithms.

\begin{table}
    \begin{minipage}{0.5\textwidth}
    \tbl{Rock-Paper-Scissors Reward Matrix}{
    \begin{tabular}{l|ccc}
    & r & p & s\\
    \hline
    r & 0,0 & -1,1 & 1,-1 \\
    p & 1,-1 & 0,0 & -1,1 \\
    s & -1,1 & 1,-1 & 0,0 \\
    \end{tabular}
    }
    \label{table:rps_reward_matrix}
    \end{minipage}
    \begin{minipage}{0.5\textwidth}
    \tbl{Biased Rock-Paper-Scissors Reward Matrix}{
        \begin{tabular}{l|ccc}
        & r & p & s\\
        \hline
        r & 50,50 & 25,75 & 100,0 \\
        p & 75,25 & 50,50 & 45,55 \\
        s & 0,100 & 55,45 & 50,50 \\
        \end{tabular}
    }
    \label{table:biased_rps_reward_matrix}
    \end{minipage}
\end{table}

To demonstrate the problem with competitive games, let's analyse some simple normal form games. Consider the simple one-shot matrix game rock-paper-scissors (RPS) in Table~\ref{table:rps_reward_matrix}. Each entry represents the reward for the row player and the column player, respectively. The only NE in the game is that two players play each of the three actions with 1/3 chance. If any player deviates from the NE, their expected reward does not increase, and the other player can exploit them by playing the action that counters their most frequent action. Table~\ref{table:biased_rps_reward_matrix} defines a more complicated game called biased rock-paper-scissors. The only NE in the game is that both players adopt the following policy.
\begin{equation}
    \pi=(P(r)=0.0625,P(p)=0.6250,P(s)=0.3125)
    \label{eq:biased_rps_policy}
\end{equation}

A well-known direct consequence of (\ref{eq:nash}) in game theory is that at any NE $\pi^n$, given the policy set of the other players $\pi^{i-1}$, player $i$ is indifferent from all actions that are chosen with a positive probability. This means that the actions are split into two sets, the worthy actions that are chosen with a positive probability and the worthless ones that will never be piked. For example, a variation of RPS could incorporate an always-losing action, where the player receives negative reward once executing it, regardless of the opponent’s action. Such an action should never be selected by the players. Additionally, the expected reward will be the same for every worthy action. In other words, the player can choose the worthy actions arbitrarily, given that other players commit to the NE. If we calculate the expected reward of each action in the biased rock-paper-scissors example where the opponent uses the NE strategy, we will see all three actions indeed have the same payoff. The above fact can be formally stated as the following. Assume $\pi^n$ is a NE. Let $Q_i(a_j, \pi^{-i})$ be the expected reward of action $a_j \in A_i$ for agent $i$, given the rest of the profile $\pi^{-i}$. Let $A^+_i$ denote the set of worthy actions of agent $i$, namely $A^+_i = \{a | a \in A_i \wedge \pi_i(a) > 0 \}$. We have:
\begin{equation}
    \begin{aligned}
    \forall a_j, a_k \in A^+_i, Q_i(a_j, \pi^{-i}) = Q_i(a_k, \pi^{-i})
    \label{eq:nash_indifferent}
    \end{aligned}
\end{equation}

Suppose a IVBL algorithm can converge to a NE, then all worthy action's Q-values must converge to the same for an agent. It does not matter how the algorithm computes a probability distribution over the actions at the end, it cannot make any discriminations among those worthy actions based on equal statistics. Such an algorithm will output at best a uniform distribution over the worthy actions at a NE. This contradicts with the assumption that they converge to a NE unless the NE is full of uniformly random policies which is not the case in most games. We have shown that pure IVBL approaches cannot work through proof by contradiction.
